Compositional Quantitative Resource Analysis of C Programs

### 2014-06-22 paper/main.tex
more work on section 2

In software engineering and software verification, we often would like
to have static information about the quantitative behavior of
programs.  For example, stack and heap-space bounds are important to
ensure the reliability of safety-critical systems~\cite{veristack14,Regehr05}.
% Clock-cycle bounds are needed to guarantee the safety of real-time
% systems [XXX wilhelm].  
Static energy usage information is critical for autonomous systems and
has applications in cloud computing~\cite{CohenZSL12,CarrollH10}.
Worst-case time bounds can help to create constant-time
implementations that prevent side-channel
attacks~\cite{KasperS09,BartheBCLP14}.  Loop and recursion-depth
bounds are used to ensure the accuracy of programs that are executed
on unreliable hardware~\cite{CarbinMR13} and complexity bounds are
needed to verify cryptography protocols~\cite{BartheGB09}.
% XXX Mention differential privacy here?
%
In general, quantitative performance information at design time can
provide useful feedback for developers.

@@

We have evaluated the quantitative logic and the automatic analysis
with system code and examples from the literature. \pref{app:cat}
contains more than $30$ (XXX) challenging loop and recursion patterns
that we collected from open source software and the literature.  Our
analysis can find asymptotically tight bounds for $30$ all but 3 of
these patterns and in most cases the derived constant factors are
tight.  We compared our automatic analyzer to other resource-analysis
systems.  We selected $5$ examples (see Section XXX) for which we
derived linear bounds and translated them to term-rewriting systems by
hand in order to run them with the tools KoAT and Pubs.  Both tools
and Rank were not able to find linear bounds for the examples.  LOOPUS
is able to derive a bound on XXX of the examples.  But in contrast to
our analysis, the constant factors are not tight.  We also contacted
the authors of SPEED but were not able to obtain a tool to perform a
comparison.

@@

\caption{Example derivation using amortized reasoning, we noted
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$, we use the
  tick metric that assigns cost $n$ to \lstinline{tick(n)} statements and
  0 to other operations.
  }
\label{fig:xmplinc}
\end{figure}

It is for instance possible to derive the following quantitative
triple that holds for all stack metrics $M$.  A stack metric assigns
cost $0$ to all evaluation steps except function calls.  Before a call
$f(x)$, we consume $M(f) > 0$ resources and after the call we return
$-M(f)$ resources.  So $M(f)$ corresponds to the stack-frame size of
the function $f$.
$$
XXX
$$
XXX explanation of triple XXX

We say that the analysis is amortized because the use of a potential
function $\Phi$ gives the flexibility to amortize some operations.
For example, if program states are simple mappings from variables to
natural numbers, we have the following judgement (provided that
initially $\state(\code x) \ge \state(\code y) \ge 0$).
$$
\htriple
{\lambda \state . \, 2 {\cdot} \state (\code x) + 10}
{\code{ x = x -  y}}
{\lambda \state . \, 2 {\cdot} \state(\code x) + 2 {\cdot} \state(\code y) + 10}
$$
This must be understood as a motion of potential from the
variable~$\code x$ to the variable~$\code y$. If some resource is
consumed $\code y$ times afterwards, we can then use $\code y$'s
potential to pay for it. However, the initial potential function
$\Phi$ only refers to the variable~$\code x$: the potential function
allows us to pass on the cost of the subsequent resource consumption
to~$\code x$.  This flexibility combined with intelligent decisions on
potential motions allows us to design an automatic analysis to
discover linear invariants for surprisingly convoluted programs.

@@

To this end, we fix the shape of the potential functions $\Phi$ so
that it becomes possible to use \emph{linear programming} (LP) to
compute a derivation in the logic.  If we assume that a program state
$\state$ simply maps variable to integers then we require
$$
\Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}} q_{(x,y)} {\cdot} |[\state(x),\state(y)]|
$$
where $q_{(x,y)} \in \Qplus$ and $|[a,b]| = \max(b-a,0)$.  Note that
$q_0$ is the constant potential and that we treat constants as program
variables.  For instance, we always have $q_{(0,x)}|[0,x]|$ as a
component of $\Phi(\state)$ if $x \in \dom{\state}$.

We then develop an inference rule for each syntactic construct that
derives sound triples $\htriple{\Phi}P{\Phi'}$ in the quantitative
logic but enables inference by using an LP solver.  This idea is best
explained by example.  Assume for the remainder of this section that
we are interested in bounding the number of \emph{ticks} in a program.
In other words, we use a resource metric that assigns cost $n$ to the
function call \code{tick(n)} and cost $0$ to all other operations.
The cost of the following example can then be bounded by $|[x,y]|$.
\begin{lstlisting}
while (x<y) {x=x+1; tick(1);} #\hfill \textnormal{(Example 1)}#
\end{lstlisting}
To derive this bound in our automatic amortized analysis, we start
with the initial potential $\Phi_0 = |[x,y]|$ (short for
$\Phi_0(\state) = |[\state(x),\state(y)]|$) which we also use as the
loop invariant.  For the loop body we then (like in Hoare logic) have
to derive a triple like
$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  However, we can
only do so if we exploit the fact that $x<y$ at the beginning of the
loop body.  The reasoning then works as follows.  We start with the
potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
assignment.  If we denote the updated version of $x$ after the
assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
the potential before and after the assignment \code{x=x+1} holds.
This means that we have the potential $|[x,y]| + 1$ before the
statement \code{tick(1)}.  Since \code{tick(1)} consumes one resource unit,
we end up with potential $|[x,y]|$ after the loop body and have
established the loop invariant again.

\begin{figure}[t]
  \centering
\begin{lstlisting}[mathescape]
${\color{blue}%
  \cdot; \, 0 + 1 {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
while (x<y) {
  ${\color{blue}%
  x{<}y; \; 0 + 1 {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
  x=x+1; 
  ${\color{blue}%
  \dashv x{\leq}y; \, 1 + 1 {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
  tick(1);
  ${\color{blue}%
  \dashv x{\leq}y; \, 0 + 1 {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
}
${\color{blue}%
 \dashv x{\geq}y; 0 + 1 {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
\end{lstlisting}
  \caption{Derivation of a bound on the number of ticks for Example~1.}
  \label{fig:ex1}
\end{figure}
More formally, we develop (in section XXX) a judgement 
$$
\Gamma; Q \vdash S \dashv Q'; \Gamma'
$$
such that $\Gamma$ contains logical assertions like $x<y$ that we
collect along the conditional branches of the program and $Q$ is a
family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
indexed by the variables that are currently in scope.  The logical
context $\Gamma$ is simply a conjunction of inequalities between.  It
is used to determine at the assignment if we can extract constant
potential to pay for future cost.  It is important to note that the
reasoning the assertions in $\Gamma$ are very basic.  We do not
perform any fixpoint computations or invariants.  \pref{fig:ex1} shows
the derivation for Example 1.  Note that the logical assertions in
$\Gamma$ are only used in the assignment \code{x=x+1}.

It would of course be possible to keep track more sophisticated
assertions in the logical context $\Gamma$.  However, this simple form
is sufficient for the examples we considered and we can efficiently
decide queries such as $\Gamma \implies x<y \; ?$, which we are
interested in.  Nevertheless, this logical part of our analysis system
is orthogonal to the quantitative part and can be easily extended if
necessary.

Note that our notion of a potential function is a refinement of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}
  while (x<y) {tick(-1); x=x+1; tick(1);} #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
notations can be seamlessly combined in a larger program.
\begin{lstlisting}
  while (x<y) {x=x+1; tick(10);} #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
%
To understand how the analysis works, it is instructive to look at
some more involved examples.  Examples \emph{speed\_1} and
\emph{speed\_2} are taken from Gulwani et al~\cite{GulwaniMC09}.
They are both handled by the SPEED tool~\cite{GulwaniMC09} but
require inference of a \emph{disjunctive invariant}.  In the abstract
interpretation community, these invariants are known to be notoriously
difficult to handle.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is more tricky and trying to understand how it
works may be challenging.  However, with the amortized analysis in
mind, using the potential transfer reasoning, it is almost trivial to
prove a bound.  While the SPEED tool has to find a fairly
involved invariant for the loop, our tool is simply reasoning locally
and works without any clever tricks. We obtain the tight bound $|[x,
n]| + |[z, n]|$.

@@

As mentioned, the analysis naturally handles advanced control flow
like \code{break} and \code{return} statements, and mutual recursion.
\pref{fig:ex_rec} contains two mutually-recursive functions with their
automatically derived bounds.  The function \code{count\_down}
decrements its first argument $x$ until it reaches the second argument
$y$. It then recursively calls the function \code{count\_up}.  The
function \code{count\_up} is dual to \code{count\_down}.  Here, we
count up $y$ by $2$ and recursively call \code{count\_down}.  Our
analysis system can compute a tight bound.  On this example, it
amounts to computing the meeting point of two trains that approach
each other with different speeds.
% This is comparable to the
% determination of the point at which to trains meet if they approach
% each other with different speeds.

>>>>

In software engineering and software verification, we often would like
to have static information about the quantitative behavior of
programs.  For example, stack and heap-space bounds are important to
ensure the reliability of safety-critical systems~\cite{veristack14,Regehr05}.
% Clock-cycle bounds are needed to guarantee the safety of real-time
% systems [XXX wilhelm].  
Static energy usage information is critical for autonomous systems and
has applications in cloud computing~\cite{CohenZSL12,CarrollH10}.
Worst-case time bounds can help to create constant-time
implementations that prevent side-channel
attacks~\cite{KasperS09,BartheBCLP14}.  Loop and recursion-depth
bounds are used to ensure the accuracy of programs that are executed
on unreliable hardware~\cite{CarbinMR13} and complexity bounds are
needed to verify cryptographic protocols~\cite{BartheGB09}.
% XXX Mention differential privacy here?
%
In general, quantitative performance information at design time can
provide useful feedback for developers.

@@

We have evaluated the quantitative logic and the automatic analysis
with system code and examples from the literature. \pref{app:cat}
contains more than $30$ (XXX) challenging loop and recursion patterns
that we collected from open source software and the literature.  Our
analysis can find asymptotically tight bounds for $30$ all but 3 of
these patterns and in most cases the derived constant factors are
tight.  We compared our automatic analyzer to other resource-analysis
systems.  We selected $5$ examples (see Section XXX) for which we
derived linear bounds and translated them to term-rewriting systems by
hand in order to run them with the tools KoAT, PUBS, and Rank.  The
tools were not able to find linear bounds for the examples.  LOOPUS is
able to derive a bound on XXX of the examples.  But in contrast to our
analysis, the constant factors are not tight.  We also contacted the
authors of SPEED but were not able to obtain a tool to perform a
comparison.

@@

\caption{Example derivation using amortized reasoning.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
  tick metric that assigns cost $n$ to \lstinline{tick(n)} statements and
  0 to other operations.
  }
\label{fig:xmplinc}
\end{figure}

It is for instance possible to derive the following quantitative
triple for a binary-search that holds for all stack metrics $M$.  A
stack metric assigns cost $0$ to all evaluation steps except function
calls.  Before a call $f(x)$, we consume $M_f > 0$ resources and
after the call we return $M_f$ resources that can be used in
subsequent function calls.  So $M_f$ corresponds to the stack-frame
size of the function $f$.
%
\begin{center}
${ (Z = \log_2(h-l)) + Z{\cdot}M_\code{bsearch} \} }$
\code{bsearch(x,l,h)}
${ \{ Z{\cdot}M_\code{bsearch} \} }$

\end{center}
%
In the pre and post condition, the logical variable $Z$ is used to
relate the size of the input with the potential that is left after the
function call.  The logical part $Z = \log_2(h-l)$ of the precondition
is $0$ if $Z = \log_2(\state_0(h)-\state_0(l))$ in the initial state
$\state_0$ and $\infty$ otherwise.

Assume for the remainder of this section that we are interested in
bounding the number of \emph{ticks} in a program.  In other words, we
use a resource metric that assigns cost $n$ to the function call
\code{tick(n)} and cost $0$ to all other operations.  The argument $n$
is an integer and a negative number means that $-n$ resources are
returned.

We say that the analysis is amortized because the use of a potential
function $\Phi$ enables us to amortize the cost of operations like in
typical textbook examples that use the potential method of Tarjan's
amortized analysis.  As an illustrative example, \pref{fig:xmplinc}
shows a program that repeatably increments a binary counter that is
implemented with a boolean-valued ($0$ or $1$) array.  The derived
bound in the precondition states that the number of ticks that are
executed by the program are bounded by $\#_1(a) + 2k$.  Here, $k$ is
the number of increments to the counter that are performed by the
program and $ \#_1(a) = \# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ is
the number of $1$'s in the array $a$.  We have inserted the statement
\code{tick(1)} into the program whenever we update the array $a$.

To derive the bound for the program in~\pref{fig:xmplinc} in the
quantitative Hoare logic, we follow the textbook
reasoning~\cite{Algorithms}.  When we assign \code{a[x]=0} in the
inner loop, we know that $a[x]=1$ before the assignment.  As a result,
we have $\#_1(a) = \#_1(a') + 1$ if $a'$ is the array $a$ after the
assignment.  So we obtain one resource unit in addition to the
invariant (i.e., $\Phi = \#_1(a) + 2k + 1$) that we can use to pay for the
cost of the statement \code{tick(1)}.  To pay for the tick statement
in the conditional, we use the fact that $k>0$ to write $2k$ as $1 +
2(k-1) +1$.  Dual to the first assignment, we have $\#_1(a) +1 =
\#_1(a')$ if $a$ is the array before and $a'$ is the array after the
assignment \code{a[x]=1}.  After paying one potential unit for the
statement $\code{tick(1)}$, we are left with the potential $\#_1(a) +
2(k - 1)$.  But since we increment $k$ at the end of the outer loop,
we can establish again our loop invariant $\#_1(a) + 2k$.

See Section XXX for more a formal definition of the logic and
additional example derivation.

% For example we have the following judgement (provided that initially
% $\state_0(\code x) \ge \state_0(\code y) \ge 0$).
% $$
% \htriple
% {2{\cdot}x + 10}
% {\code{ x = x -  y}}
% {2{\cdot}x + 2{\cdot}y + 10}
% $$
% This must be understood as a motion of potential from the
% variable~$\code x$ to the variable~$\code y$. If some resource is
% consumed $\code y$ times afterwards, we can then use $\code y$'s
% potential to pay for it. However, the initial potential function
% $\Phi$ only refers to the variable~$\code x$: the potential function
% allows us to pass on the cost of the subsequent resource consumption
% to~$\code x$.  This flexibility combined with intelligent decisions on
% potential motions allows us to design an automatic analysis to
% discover linear invariants for surprisingly convoluted programs.

@@

To this end, we fix the shape of the potential functions $\Phi$ so
that it becomes possible to use \emph{linear programming} (LP) to
compute a derivation in the logic.  If we assume that a program state
$\state$ simply maps variable to integers then we require
$$
\Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}} q_{(x,y)} {\cdot} |[\state(x),\state(y)]|
$$
where $q_{(x,y)} \in \Qplus$ and $|[a,b]| = \max(b-a,0)$.  Note that
$q_0$ is the constant potential and that we treat constants as program
variables.  For instance, we always have $q_{(0,x)}|[0,x]|$ as a
component of $\Phi(\state)$ if $x \in \dom{\state}$.
%
We then develop an inference rule for each syntactic construct that
derives sound triples $\htriple{\Phi}P{\Phi'}$ in the quantitative
logic but enables inference by using an LP solver.

This idea is best explained by example.  If we again use the
\emph{ticks} metric that assigns cost $n$ to the function call
\code{tick(n)} and cost $0$ to all other operations then the cost of
the following example can be bounded by $|[x,y]|$.
\begin{lstlisting}
while (x<y) {x=x+1; tick(1);} #\hfill \textnormal{(Example 1)}#
\end{lstlisting}
To derive this bound in our automatic amortized analysis, we start
with the initial potential $\Phi_0 = |[x,y]|$ (short for
$\Phi_0(\state) = |[\state(x),\state(y)]|$) which we also use as the
loop invariant.  For the loop body we then (like in Hoare logic) have
to derive a triple like
$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  However, we can
only do so if we utilize the fact that $x<y$ at the beginning of the
loop body.  The reasoning then works as follows.  We start with the
potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
assignment.  If we denote the updated version of $x$ after the
assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
the potential before and after the assignment \code{x=x+1} holds.
This means that we have the potential $|[x,y]| + 1$ before the
statement \code{tick(1)}.  Since \code{tick(1)} consumes one resource unit,
we end up with potential $|[x,y]|$ after the loop body and have
established the loop invariant again.

\begin{figure}[t]
  \centering
\begin{lstlisting}[mathescape]
${\color{blue}%
  \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
while (y-x>K) {
  ${\color{blue}%
  x{<}y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
  x=x+K; 
  ${\color{blue}%
  \dashv x{\leq}y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
  tick(T);
  ${\color{blue}%
  \dashv x{\leq}y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
}
${\color{blue}%
 \dashv x{\geq}y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
\end{lstlisting}
  \caption{Automatic derivation of a tight bound on the number of
    ticks for a standard \emph{for loop}.  The parameters $K>0$ and
    $T>0$ are not program variables but denote a concrete constants in
    the program.  The reasoning works uniformly for all such $K$ and
    $T$.}
  \label{fig:ex1}
\end{figure}
More formally, we develop (in section XXX) a judgement 
$$
\Gamma; Q \vdash S \dashv Q'; \Gamma'
$$
such that $\Gamma$ contains logical assertions like $x<y$ that we
collect along the conditional branches of the program and $Q$ is a
family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
indexed by the variables that are currently in scope.  The logical
context $\Gamma$ is simply a conjunction of inequalities between
linear combinations variables.  It is used to determine at the
assignment if we can extract constant potential to pay for future
cost.  It is important to note that the reasoning about assertions in
$\Gamma$ are very basic.  We do not perform any fixpoint computations
and only derive trivial loop invariants about variables that are
unchanged in the loop body.

Note that our notion of a potential function is a refinement of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}
  while (x<y) {tick(-1); x=x+1; tick(1);} #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
tick annotations can be seamlessly combined in a larger program.
\begin{lstlisting}
  while (x<y) {x=x+1; tick(10);} #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
%
\pref{fig:ex1} shows a derivation of the bound
$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
version of Example~1 in which increment $x$ be a constant $K>0$ and
consume $T>0$ resources in each iteration.  The reasoning is similar
to the description of Example~1 except that we obtain the potential
$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
assertions in $\Gamma$ are only used in rule for the assignment
\code{x=x+1}.  To the best of our knowledge, no other tool is
currently capable of deriving a tight bound on the cost of such a
loop.  For $T=1$ (many systems focus on the number of loop iterations
without a cost model) and $K=10$, KoAT~\cite{BrockschmidtEFFG14}
computes the bound XXX, PUBS~\cite{AlbertAGPZ12} computes the bound
XXX, Rank~\cite{AliasDFG10} computes the bound XXX, and
LOOPUS~\cite{SinnZV14} computes the bound XXX.

It might be surprising that we only track simple linear relations in
the logical context $\Gamma$.  It would of course be possible to keep
track more sophisticated assertions.  However, this simple form is
sufficient for the examples we considered and we can efficiently
decide queries such as $\Gamma \implies x<y \; ?$ that we make in the
assignment rule.  Nevertheless, this logical part of our analysis
system is orthogonal to the quantitative part and can be easily
extended if necessary.

To understand the capabilities of the automatic analysis, it is instructive to
look at some more involved examples.  Examples \emph{speed\_1} and
\emph{speed\_2} are taken from Gulwani et al~\cite{GulwaniMC09}.  They
are both handled by the SPEED tool~\cite{GulwaniMC09} but require
inference of a \emph{disjunctive invariant}.  In the abstract
interpretation community, these invariants are known to be notoriously
difficult to handle.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is more tricky and trying to understand how it
works may be challenging.  However, with the amortized analysis in
mind, using the potential transfer reasoning, it is almost trivial to
prove a bound.  While the SPEED tool has to find a fairly
involved invariant for the loop, our tool is simply reasoning locally
and works without any clever tricks. We obtain the tight bound $|[x,
n]| + |[z, n]|$.

@@

As mentioned, the analysis naturally handles advanced control flow
like \code{break} and \code{return} statements, and mutual recursion.
\pref{fig:ex_rec} contains two mutually-recursive functions with their
automatically derived tick bounds.  The function \code{count\_down}
decrements its first argument $x$ until it reaches the second argument
$y$. It then recursively calls the function \code{count\_up}.  The
function \code{count\_up} is dual to \code{count\_down}.  Here, we
count up $y$ by $2$ and recursively call \code{count\_down}.  Our
analysis system can compute a tight bound.  On this example, it
amounts to computing the meeting point of two trains that approach
each other with different speeds.
% This is comparable to the
% determination of the point at which to trains meet if they approach
% each other with different speeds.
